Evolution of pre-copulatory and post-copulatory strategies of inbreeding avoidance

### 2017-07-20 ms_pre_post.Rmd
Some finishing initial edits

When post-copulatory inbreeding avoidance alleles ($F_{a}$) had no effect (i.e., $F_{g}$ values were fixed to zero), meaning that $F_{p}$ could not evolve, $P_{a}$ alleles underlying polyandry always decreased to negative values over generations (red lines, Figure 1A,C,E,G). This shows that despite strong inbreeding depression in offspring viability, there is selection against unconditional polyandry even given zero direct cost ($c_{P}=0$; Figure 1A). This is because $M_{a}$ values became negative over generations, meaning that females typically avoided inbreeding through their initial mating (blue lines, Figure 1A,C,E,G). Polyandrous females that subsequently sampled more males from the available population were consequently more likely to mate with some relatives and hence produce some inbred offspring with low viability [@Duthie].

@@

However, when both pre-copulatory and post-copulatory inbreeding avoidance could evolve, their relative evolutionary dynamics depended on their relative costs. When pre-copulatory but not post-copulatory inbreeding avoidance was costly ($c_{M}=0.02$ and $c_{F}=0$), pre-copulatory inbreeding avoidance initially evolved (i.e., $M_{a} < 0$) but then evolved back towards random mating (i.e., $M_{a} \approx 0$) following increasing evolution of post-copulatory inbreeding avoidance and polyandry (Figure 3B). Similarly, when post-copulatory but not pre-copulatory inbreeding avoidance was costly ($c_{F}=0.02$ and $c_{M}=0$), post-copulatory inbreeding avoidance initially evolved (i.e., $F_{a} < 0$) before evolving back to random fertilisation (i.e., $F_{a} \approx 0$) after ca $20000$ generations (Figure 3D).

When pre-copulatory inbreeding avoidance was costly, allowing evolution of cost free post-copulatory inbreeding avoidance greatly facilitated evolution of polyandry (Figure 3A versus 3B). However when post-copulatory inbreeding avoidance was costly, allowing evolution of cost free pre-copulatory inbreeding avoidance caused $P_{a}$ alleles to decrease to very negative values, reducing expression of polyandry (Figure 3C versus 3D; polyandry was cost free in all these simulations). Results for all possible cost combinations of $0$ and $0.02$ are provided in Supporting Information (Figure S9).

@@

When polyandry alleles ($P_{a}$) were fixed to be positive so that all females were expected to mate multiply and post-copulatory inbreeding allele ($F_{a}$) values were fixed for adaptive inbreeding avoidance, pre-copulatory inbreeding avoidance evolved (i.e., $M_{a}$ values became increasingly negative; Figure 4A). Such evolution still occurred, but to a much smaller degree, when pre-copulatory inbreeding avoidance was costly (Figure 4B). However, after $40000$ generations, $M_{a}$ values were less negative when post-copulatory inbreeding avoidance and polyandry were fixed at non-zero values than when they also evolved from initial values of zero ($-15.45$ vs. $-17.62$; compare the blue lines in Figures 4A versus Figure 1B). This shows that fixation of post-copulatory inbreeding avoidance weakens selection for pre-copulatory inbreeding avoidance.

Conversely, when pre-copulatory inbreeding allele ($M_{a}$) values were fixed for adaptive inbreeding avoidance, mean $F_{a}$ allele values did not consistently become negative over generations (Figure 4C,D). Fixation of pre-copulatory inbreeding avoidance therefore prevented evolution of post-copulatory inbreeding avoidance. In this case, mean $P_{a}$ allele values consistently decreased over generations, reflecting selection against polyandry regardless of whether or not post-copulatory inbreeding avoidance was costly (Figure 4C,D). Consequently, when $c_{F}=0$, $F_{a}$ allele values had no effect because females were almost exclusively monandrous, resulting in high drift of $F_{a}$ values (resulting in variation among replicates illustrated by the wide standard errors in Figure 4C). However, when $c_{F}=0.02$, $F_{a}$ values remained near zero to minimise direct costs. This lack of selection for post-copulatory inbreeding avoidance was driven by a lack of polyandry, and therefore an inability of females to bias fertilisation among multiple mates. When pre-copulatory inbreeding avoidance and polyandry were both fixed ($M_{a}=-10$ and $P_{a}=1$), $F_{a}$ allele values evolved to similarly negative means as $M_{a}$ allele values in Figure 4A,B (see Supporting Information Figure S10).

@@

Evolution of both pre-copulatory and post-copulatory inbreeding avoidance occurred in our model, but were affected by the evolution of polyandry and the existence of asymmetric costs. One cost free strategy of inbreeding avoidance precluded another more costly strategy from persisting in a focal population (Figure 3). Consequently, long-term persistence of both pre-copulatory and post-copulatory inbreeding avoidance might not be expected in populations given sufficient cost asymmetry. Since direct costs strongly affected evolution of the focal phenotype(s), quantifying costs is necessary for predicting evolution of inbreeding strategies.

**To facilitate a general comparison of the effects of direct costs across phenotypes, all costs in our model were manifested as an increased probability of total reproductive failure. We thereby interpreted as a cost of polyandy as an elevated risk of predation as might be incurred while searching for mates [e.g., @Koga1998], a cost of pre-copulatory inbreeding strategy as a risk associated with male harrassment [e.g., @Rowe1994], and a cost of post-copulatory inbreeding strategy as an up-front depletion of energy needed for post-copulatory kin discrimination [e.g., @Gasparini2011]**. While these costs reflect empirical observations in some populations and are therefore biologically realistic, **more mechanistic** costs could be **assumed or predicted from future models tailored to specific empirical systems.** In empirical systems, quantifying direct costs of phenotypes associated with polyandry and mate choice is challenging [@Pomiankowski1987; @Kokko2003]. @Pomiankowski1987 categorised four types of costs that are relevant to **mating frequency and** mate choice, which include costs of elevated risks of predation or disease transmission, and costs incurred through time or energy expenditure. For example, highly polyandrous females might experience increased risk of disease transmission [@Roberts2015a]. Such a cost would more realistically apply to a female's realised number of mates rather than her tendency for polyandry. Polyandrous females might also risk harm caused by sexual conflict over multiple mating [e.g., @Arnqvist2005a; @Parker2006]. Inbreeding theory predicts that males should be more tolerant of inbreeding than females, leading to sexual conflict over inbreeding [@Parker1979; @Parker2006; @Kokko2006; @Duthie2015a]. Future models could **therefore** formally consider sexual conflict over both polyandry and inbreeding. Overall, such different types of **potentially interacting** costs might affect selection both directly and indirectly through interactions between reproductive strategies, thereby affecting evolution of entire reproductive systems. Evolution of reproductive strategies therefore needs to be considered in the broader context of potentially interacting and phenotypes and their costs.

@@

Post-copulatory inbreeding avoidance cannot be effectively realised if females are not polyandrous in any form, and is likely to be most effective for highly polyandrous females that can choose amongst sperm contributed by multiple mates and therefore bias fertilisation accordingly. In contrast, pre-copulatory inbreeding avoidance mechanisms are most critical for females that mate only once and therefore have no post-copulatory opportunity to avoiding inbreeding. This theory is borne out in our simulation results, as selection for, and consequent evolution of, post-copulatory inbreeding avoidance was negligible in populations where polyandry did not evolve, resulting in high drift of allele values over generations due to the inability of females to express post-copulatory choice (e.g., Figure 4C,D). Evolution of pre-copulatory inbreeding avoidance was also typically slightly stronger when polyandry did not evolve (e.g., Figure 3A vs 3B; see also Supporting Information Figure S9). In addition to initial polyandry causing evolution of post-copulatory inbreeding avoidance, polyandry might also covary positively with post-copulatory inbreeding avoidance due to the feed back effect that post-copulatory inbreeding avoidance has on facilitating evolution of polyandry itself, as observed in our model (Figure 1). It would therefore be interesting to test the hypothesis that, across taxa, the occurrence of post-copulatory inbreeding avoidance covaries positively, and the occurrence of pre-copulatory inbreeding avoidance covaries negatively, with the degree of polyandry. To test this hypothesis, more work is needed to quantify the degree to which females of different species engage in polyandry and the degree to which females express both pre-copulatory and post-copulatory inbreeding avoidance.

>>>>

When post-copulatory inbreeding avoidance alleles ($F_{a}$) had no effect (i.e., $F_{g}$ values were fixed to zero), meaning that $F_{p}$ could not evolve, $P_{a}$ alleles underlying polyandry decreased to negative values over generations (red lines, Figure 1A,C,E,G). This shows that despite strong inbreeding depression in offspring viability, there is selection against unconditional polyandry even given zero direct cost ($c_{P}=0$; Figure 1A). This is because $M_{a}$ values became negative over generations, meaning that females typically avoided inbreeding through their initial mating (blue lines, Figure 1A,C,E,G). Polyandrous females that subsequently sampled more males from the available population were consequently more likely to mate with some relatives and hence produce some inbred offspring with low viability [@Duthie].

@@

However, when both pre-copulatory and post-copulatory inbreeding avoidance could evolve, their relative evolutionary dynamics depended on their relative costs. When pre-copulatory but not post-copulatory inbreeding avoidance was costly ($c_{M}=0.02$ and $c_{F}=0$), pre-copulatory inbreeding avoidance initially evolved (i.e., $M_{a} < 0$) but then evolved back towards random mating (i.e., $M_{a} \approx 0$) following increasing evolution of post-copulatory inbreeding avoidance and polyandry (Figure 3B). Similarly, when post-copulatory but not pre-copulatory inbreeding avoidance was costly ($c_{F}=0.02$ and $c_{M}=0$), post-copulatory inbreeding avoidance initially evolved (i.e., $F_{a} < 0$) before evolving back to random fertilisation (i.e., $F_{a} \approx 0$) after ca $20000$ generations (Figure 3D). **This evolution of a high cost strategy back toward random mating or fertilisation in the presence of a relatively low cost alternative inbreeding strategy is a general consequence of differential selection on each strategy induced by cost asymmetry. Hence, this pattern can be observed for both non-zero low relative costs (e.g., values of 0.01 and 0.03; Figure S9) and smaller cost asymmetries (e.g., values of 0.01 and 0.02; Figure S10).**

When pre-copulatory inbreeding avoidance was costly, allowing evolution of cost free post-copulatory inbreeding avoidance greatly facilitated evolution of polyandry (Figure 3A versus 3B). However when post-copulatory inbreeding avoidance was costly, allowing evolution of cost free pre-copulatory inbreeding avoidance caused $P_{a}$ alleles to decrease to very negative values, reducing expression of polyandry (Figure 3C versus 3D; polyandry was cost free in all these simulations). Results for all possible cost combinations of $0$ and $0.02$ are provided in Supporting Information (Figure S**11**).

@@

When polyandry alleles ($P_{a}$) were fixed to be positive so that all females were expected to mate multiply and post-copulatory inbreeding allele ($F_{a}$) values were fixed for adaptive inbreeding avoidance, pre-copulatory inbreeding avoidance evolved (i.e., $M_{a}$ values became increasingly negative; Figure 4A). Such evolution still occurred, but to a much smaller degree, when pre-copulatory inbreeding avoidance was costly (Figure 4B). However, after $40000$ generations, $M_{a}$ values were less negative when post-copulatory inbreeding avoidance and polyandry were fixed at non-zero values than when they also evolved from initial values of zero ($-15.45$ vs. $-17.62$; compare the blue lines in Figures 4A versus Figure 1B). **Hence, we hypothesise that existence of fixed post-copulatory inbreeding avoidance can weaken selection and subsequent evolution of pre-copulatory inbreeding avoidance.**

Conversely, when pre-copulatory inbreeding allele ($M_{a}$) values were fixed for adaptive inbreeding avoidance, mean $F_{a}$ allele values did not consistently become negative over generations (Figure 4C,D). **Hence, we hypothesise that existence of fixed pre-copulatory inbreeding avoidance can prevent evolution of post-copulatory inbreeding avoidance.** **In our model**, mean $P_{a}$ allele values consistently decreased over generations, reflecting selection against polyandry regardless of whether or not post-copulatory inbreeding avoidance was costly (Figure 4C,D). Consequently, when $c_{F}=0$, $F_{a}$ allele values had no effect because females were almost exclusively monandrous, resulting in high drift of $F_{a}$ values (resulting in variation among replicates illustrated by the wide standard errors in Figure 4C). However, when $c_{F}=0.02$, $F_{a}$ values remained near zero to minimise direct costs. This lack of selection for post-copulatory inbreeding avoidance was driven by a lack of polyandry, and therefore an inability of females to bias fertilisation among multiple mates. When pre-copulatory inbreeding avoidance and polyandry were both fixed ($M_{a}=-10$ and $P_{a}=1$), $F_{a}$ allele values evolved to similarly negative means as $M_{a}$ allele values in Figure 4A,B (see Supporting Information Figure S**12**).

@@

Evolution of both pre-copulatory and post-copulatory inbreeding avoidance occurred in our model, but were affected by the evolution of polyandry and the existence of asymmetric costs. One cost free strategy of inbreeding avoidance precluded another more costly strategy from persisting in a focal population (Figure 3). **Hence, our model demonstrates a novel mechanism by which degenerate inbreeding avoidance strategies can interact with one another, and this mechanism might usefult to consider when developing hypotheses concerning reproductive strategy within and across systems.** **In particular**, **indefinite** persistence of both pre-copulatory and post-copulatory inbreeding avoidance **should not be** expected in populations given **a** sufficient**ly large and sustained cost asymmetry.** **However, the length of time required for the more costly inbreeding strategy to go extinct might be quite long (Figure 3), and spatial or temporal variation in costs might facilitate the coexistence of multiple inbreeding avoidance strategies. Nevertheless,** since direct costs strongly affected evolution of the focal phenotype(s), quantifying costs is necessary for predicting evolution of inbreeding strategies.

**To facilitate a general comparison of the effects of direct costs across phenotypes, all costs in our model were manifested as an increased probability of total reproductive failure. We thereby interpreted as a cost of polyandy as an elevated risk of predation as might be incurred while searching for mates [e.g., @Koga1998], a cost of pre-copulatory inbreeding strategy as a risk associated with male harrassment [e.g., @Rowe1994], and a cost of post-copulatory inbreeding strategy as an up-front depletion of energy needed for post-copulatory kin discrimination [e.g., @Gasparini2011]**. While these costs reflect empirical observations in some populations and are therefore biologically realistic, **more mechanistic** costs could be **assumed or predicted from future models tailored to specific empirical systems.** In empirical systems, quantifying direct costs of phenotypes associated with polyandry and mate choice is challenging [@Pomiankowski1987; @Kokko2003]. @Pomiankowski1987 categorised four types of costs that are relevant to **mating frequency and** mate choice, which include costs of elevated risks of predation or disease transmission, and costs incurred through time or energy expenditure. For example, highly polyandrous females might experience increased risk of disease transmission [@Roberts2015a]. Such a cost would more realistically apply to a female's realised number of mates rather than her tendency for polyandry. Polyandrous females might also risk harm caused by sexual conflict over multiple mating [e.g., @Arnqvist2005a; @Parker2006]. Inbreeding theory predicts that males should be more tolerant of inbreeding than females, leading to sexual conflict over inbreeding [@Parker1979; @Parker2006; @Kokko2006; @Duthie2015a]. Future models could **therefore more mechanistically** consider sexual conflict over both polyandry and inbreeding. Overall, such different types of **potentially interacting** costs might affect selection both directly and indirectly through interactions between reproductive strategies, thereby affecting evolution of entire reproductive systems. Evolution of reproductive strategies therefore needs to be considered in the broader context of potentially interacting and phenotypes and their costs.

@@

Post-copulatory inbreeding avoidance cannot be effectively realised if females are not polyandrous in any form, and is likely to be most effective for highly polyandrous females that can choose amongst sperm contributed by multiple mates and therefore bias fertilisation accordingly. In contrast, pre-copulatory inbreeding avoidance mechanisms are most critical for females that mate only once and therefore have no post-copulatory opportunity to avoiding inbreeding. This theory is borne out in our simulation results, as selection for, and consequent evolution of, post-copulatory inbreeding avoidance was negligible in populations where polyandry did not evolve, resulting in high drift of allele values over generations due to the inability of females to express post-copulatory choice (e.g., Figure 4C). Evolution of pre-copulatory inbreeding avoidance was also typically slightly stronger when polyandry did not evolve (e.g., Figure 3A vs 3B; see also Supporting Information Figure S**11**). In addition to initial polyandry causing evolution of post-copulatory inbreeding avoidance, polyandry might also covary positively with post-copulatory inbreeding avoidance due to the feed back effect that post-copulatory inbreeding avoidance has on facilitating evolution of polyandry itself, as observed in our model (Figure 1). It would therefore be interesting to test the hypothesis that, across taxa, the occurrence of post-copulatory inbreeding avoidance covaries positively, and the occurrence of pre-copulatory inbreeding avoidance covaries negatively, with the degree of polyandry. To test this hypothesis, more work is needed to quantify the degree to which females of different species engage in polyandry and the degree to which females express both pre-copulatory and post-copulatory inbreeding avoidance.
